---
title: "STAT 252 — In-Class Experiments Report"
author: "Cal Poly San Luis Obispo"
date: today
format:
  docx:
    toc: true
    toc-depth: 2
    number-sections: true
    fig-width: 6
    fig-height: 3.8
execute:
  echo: false
  warning: false
  message: false
params:
  sheet_id: "PASTE_YOUR_GOOGLE_SHEET_ID_HERE"
  session: "spring2026-mwf10"
  show_reaction: true
  show_memory: true
  show_framing: true
---

```{r setup}
# ---- Packages ----
library(googlesheets4)
library(dplyr)
library(ggplot2)
library(knitr)
library(scales)

# ---- Auth ----
# First time you render: a browser window will open. Sign in with the Google
# account that owns the sheet, allow access. After that, the token is cached
# and you won't be prompted again.
gs4_auth(email = TRUE)  # uses the most recently authenticated email

# ---- Helpers ----
read_tab <- function(tab) {
  tryCatch(
    read_sheet(params$sheet_id, sheet = tab) |>
      filter(session == params$session),
    error = function(e) NULL
  )
}

fmt_p <- function(p) {
  if (is.na(p)) return("NA")
  if (p < 0.001) return("< 0.001")
  formatC(p, format = "f", digits = 4)
}

# Section heading helper
section_check <- function(df, name) {
  if (is.null(df) || nrow(df) == 0) {
    cat(sprintf("> *No data found for the '%s' experiment in session `%s`. Skipping.*\n\n",
                name, params$session))
    return(FALSE)
  }
  TRUE
}
```

```{r session-info}
cat(sprintf("**Session code:** `%s`  \n", params$session))
cat(sprintf("**Report generated:** %s\n", format(Sys.time(), "%B %d, %Y at %I:%M %p")))
```

# Experiment 1 — Two Population Means: Reaction Time

```{r exp1-load}
exp1 <- if (params$show_reaction) read_tab("reaction") else NULL
exp1_ok <- section_check(exp1, "reaction")
```

```{r exp1-content, eval=exp1_ok, results='asis'}
cat("## Design\n\n")
cat("Each student was randomly assigned to one of two conditions: completing a 5-trial reaction-time task with **music** playing through headphones, or in **silence**. Their average reaction time across the 5 trials (in milliseconds) is the response.\n\n")
cat("- **Population**: All Cal Poly students who could plausibly take this course.\n")
cat("- **Sample**: The students in this class section who participated.\n")
cat("- **Observational unit**: An individual student.\n")
cat("- **Explanatory variable**: Listening condition (music vs. silence) — categorical, two levels.\n")
cat("- **Response variable**: Average reaction time across 5 trials, in ms — quantitative.\n")
cat("- **Parameter of interest**: $\\mu_{\\text{music}} - \\mu_{\\text{silence}}$, the difference in true mean reaction time between the two populations.\n\n")
```

```{r exp1-summary, eval=exp1_ok}
exp1_clean <- exp1 |>
  mutate(group = factor(group, levels = c("music", "silence")),
         avg_rt_ms = as.numeric(avg_rt_ms))

exp1_summary <- exp1_clean |>
  group_by(group) |>
  summarise(
    n = n(),
    mean_rt = round(mean(avg_rt_ms, na.rm = TRUE), 1),
    sd_rt = round(sd(avg_rt_ms, na.rm = TRUE), 1),
    median_rt = round(median(avg_rt_ms, na.rm = TRUE), 1),
    .groups = "drop"
  )

kable(exp1_summary,
      col.names = c("Group", "n", "Mean RT (ms)", "SD (ms)", "Median (ms)"),
      caption = "Summary statistics by group")
```

```{r exp1-plot, eval=exp1_ok}
ggplot(exp1_clean, aes(x = group, y = avg_rt_ms, fill = group)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA, width = 0.5) +
  geom_jitter(width = 0.12, size = 2.2, alpha = 0.75) +
  scale_fill_manual(values = c("music" = "#2a6f97", "silence" = "#c0392b")) +
  labs(x = "Listening condition", y = "Average reaction time (ms)",
       title = "Reaction time by listening condition") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

```{r exp1-test, eval=exp1_ok, results='asis'}
test1 <- t.test(avg_rt_ms ~ group, data = exp1_clean)

cat("## Hypotheses\n\n")
cat("$$H_0:\\ \\mu_{\\text{music}} - \\mu_{\\text{silence}} = 0 \\qquad H_a:\\ \\mu_{\\text{music}} - \\mu_{\\text{silence}} \\neq 0$$\n\n")

cat("## Point estimate\n\n")
pe <- diff(rev(test1$estimate))
cat(sprintf("$\\bar{x}_{\\text{music}} - \\bar{x}_{\\text{silence}} = %.1f$ ms\n\n", -pe))

cat("## Test results\n\n")
res1 <- data.frame(
  Quantity = c("Test statistic (t)", "Degrees of freedom", "p-value",
               "95% CI lower", "95% CI upper"),
  Value = c(round(test1$statistic, 3),
            round(test1$parameter, 2),
            fmt_p(test1$p.value),
            round(test1$conf.int[1], 2),
            round(test1$conf.int[2], 2))
)
print(kable(res1))

cat("\n\n## Interpretation\n\n")
cat(sprintf("- The test statistic %.3f tells us how many standard errors the observed difference in sample means lies from zero.\n", test1$statistic))
cat(sprintf("- The p-value of %s is the probability of seeing a difference at least this extreme if there were truly no effect of music on reaction time.\n", fmt_p(test1$p.value)))
ci_includes_zero <- test1$conf.int[1] <= 0 & test1$conf.int[2] >= 0
cat(sprintf("- We are 95%% confident that the true mean difference (music − silence) lies between %.1f ms and %.1f ms. ",
            test1$conf.int[1], test1$conf.int[2]))
if (ci_includes_zero) {
  cat("Because **0 falls inside** this interval, the data are consistent with no real difference in mean reaction time between the two conditions.\n\n")
} else {
  cat("Because **0 does not fall inside** this interval, we have evidence that the two conditions produce different mean reaction times.\n\n")
}
```

\newpage

# Experiment 2 — Paired t: Memory Recall Before vs. After Chunking

```{r exp2-load}
exp2 <- if (params$show_memory) read_tab("memory") else NULL
exp2_ok <- section_check(exp2, "memory")
```

```{r exp2-content, eval=exp2_ok, results='asis'}
cat("## Design\n\n")
cat("Each student was shown a string of 14 random digits for 6 seconds and asked to recall as many as possible in order (Round 1, *before*). They were then taught the **chunking** trick — grouping digits into sets of 3–4 — and shown a new 14-digit string, this time displayed pre-chunked on screen, under the same time constraint (Round 2, *after*). The response is each student's score (number of digits recalled in correct position out of 14) in each round.\n\n")
cat("- **Population**: All Cal Poly students who could plausibly take this course.\n")
cat("- **Sample**: The students in this class section who participated.\n")
cat("- **Observational unit**: An individual student (each student contributes a *pair* of scores).\n")
cat("- **Explanatory variable**: Round (before vs. after the chunking trick) — categorical, two levels, applied to the same student.\n")
cat("- **Response variable**: Number of digits recalled correctly out of 14 — quantitative.\n")
cat("- **Parameter of interest**: $\\mu_d$, the true mean of the within-student difference (after − before).\n\n")
cat("Because each student contributes both measurements, we use a **paired t-test** on the differences rather than a two-sample test.\n\n")
```

```{r exp2-summary, eval=exp2_ok}
exp2_clean <- exp2 |>
  mutate(score_before = as.numeric(score_before),
         score_after = as.numeric(score_after),
         difference = score_after - score_before)

exp2_summary <- data.frame(
  Quantity = c("n (students)", "Mean (before)", "Mean (after)",
               "Mean difference (after − before)", "SD of differences"),
  Value = c(nrow(exp2_clean),
            round(mean(exp2_clean$score_before), 2),
            round(mean(exp2_clean$score_after), 2),
            round(mean(exp2_clean$difference), 2),
            round(sd(exp2_clean$difference), 2))
)
kable(exp2_summary, caption = "Summary statistics")
```

```{r exp2-plot, eval=exp2_ok}
exp2_long <- exp2_clean |>
  mutate(student = row_number()) |>
  tidyr::pivot_longer(cols = c(score_before, score_after),
                      names_to = "round", values_to = "score") |>
  mutate(round = factor(round,
                        levels = c("score_before", "score_after"),
                        labels = c("Before", "After")))

ggplot(exp2_long, aes(x = round, y = score, group = student)) +
  geom_line(alpha = 0.4, color = "#6b6b6b") +
  geom_point(aes(color = round), size = 2.5) +
  scale_color_manual(values = c("Before" = "#c0392b", "After" = "#2d8a4e")) +
  labs(x = "", y = "Digits recalled correctly (out of 14)",
       title = "Within-student change after learning chunking") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

```{r exp2-test, eval=exp2_ok, results='asis'}
test2 <- t.test(exp2_clean$score_after, exp2_clean$score_before, paired = TRUE)

cat("## Hypotheses\n\n")
cat("$$H_0:\\ \\mu_d = 0 \\qquad H_a:\\ \\mu_d \\neq 0$$\n\n")
cat("where $\\mu_d$ is the true mean difference (after − before).\n\n")

cat("## Point estimate\n\n")
cat(sprintf("$\\bar{d} = %.2f$ digits\n\n", mean(exp2_clean$difference)))

cat("## Test results\n\n")
res2 <- data.frame(
  Quantity = c("Test statistic (t)", "Degrees of freedom", "p-value",
               "95% CI lower", "95% CI upper"),
  Value = c(round(test2$statistic, 3),
            round(test2$parameter, 2),
            fmt_p(test2$p.value),
            round(test2$conf.int[1], 3),
            round(test2$conf.int[2], 3))
)
print(kable(res2))

cat("\n\n## Interpretation\n\n")
cat(sprintf("- The test statistic %.3f tells us how many standard errors the mean within-student change lies from zero.\n", test2$statistic))
cat(sprintf("- The p-value of %s is the probability of seeing a mean change at least this large in magnitude if chunking truly had no effect on recall.\n", fmt_p(test2$p.value)))
ci_includes_zero2 <- test2$conf.int[1] <= 0 & test2$conf.int[2] >= 0
cat(sprintf("- We are 95%% confident that the true mean change in score (after − before) lies between %.2f and %.2f digits. ",
            test2$conf.int[1], test2$conf.int[2]))
if (ci_includes_zero2) {
  cat("Because **0 falls inside** this interval, the data are consistent with chunking having no effect on recall.\n\n")
} else {
  cat("Because **0 does not fall inside** this interval, we have evidence that chunking changes mean recall scores.\n\n")
}
```

\newpage

# Experiment 3 — Two Population Proportions: Framing Effect

```{r exp3-load}
exp3 <- if (params$show_framing) read_tab("framing") else NULL
exp3_ok <- section_check(exp3, "framing")
```

```{r exp3-content, eval=exp3_ok, results='asis'}
cat("## Design\n\n")
cat("Each student was randomly assigned to read one of two versions of an identical medical scenario. The **positive frame** described the surgery as having a 90% one-year survival rate. The **negative frame** described the same surgery as having a 10% one-year mortality rate. Each student then chose whether to recommend the treatment (yes / no).\n\n")
cat("- **Population**: All Cal Poly students who could plausibly take this course.\n")
cat("- **Sample**: The students in this class section who participated.\n")
cat("- **Observational unit**: An individual student.\n")
cat("- **Explanatory variable**: Frame (positive vs. negative) — categorical, two levels.\n")
cat("- **Response variable**: Whether the student recommended the treatment (yes/no) — categorical, binary.\n")
cat("- **Parameter of interest**: $p_{\\text{positive}} - p_{\\text{negative}}$, the difference in the true proportion of students who would recommend treatment under each frame.\n\n")
```

```{r exp3-summary, eval=exp3_ok}
exp3_clean <- exp3 |>
  mutate(group = factor(group, levels = c("positive", "negative")),
         choice = factor(choice, levels = c("yes", "no")))

exp3_summary <- exp3_clean |>
  group_by(group) |>
  summarise(
    n = n(),
    yes_count = sum(choice == "yes"),
    prop_yes = round(mean(choice == "yes"), 3),
    .groups = "drop"
  )
kable(exp3_summary,
      col.names = c("Frame", "n", "# Recommended", "Proportion 'yes'"),
      caption = "Recommendation rates by framing condition")
```

```{r exp3-plot, eval=exp3_ok}
ggplot(exp3_summary, aes(x = group, y = prop_yes, fill = group)) +
  geom_col(width = 0.55, alpha = 0.85) +
  geom_text(aes(label = paste0(round(prop_yes * 100), "%")),
            vjust = -0.4, size = 4.5) +
  scale_y_continuous(limits = c(0, 1.05), labels = percent_format()) +
  scale_fill_manual(values = c("positive" = "#2a6f97", "negative" = "#c0392b")) +
  labs(x = "Framing condition", y = "Proportion who recommended treatment",
       title = "Recommendation rate by frame") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

```{r exp3-test, eval=exp3_ok, results='asis'}
n_pos <- sum(exp3_clean$group == "positive")
n_neg <- sum(exp3_clean$group == "negative")
yes_pos <- sum(exp3_clean$group == "positive" & exp3_clean$choice == "yes")
yes_neg <- sum(exp3_clean$group == "negative" & exp3_clean$choice == "yes")

test3 <- prop.test(x = c(yes_pos, yes_neg), n = c(n_pos, n_neg), correct = FALSE)

# Two-proportion z statistic (manual, since prop.test gives chi-square)
p_pos <- yes_pos / n_pos
p_neg <- yes_neg / n_neg
p_pool <- (yes_pos + yes_neg) / (n_pos + n_neg)
se_pool <- sqrt(p_pool * (1 - p_pool) * (1/n_pos + 1/n_neg))
z_stat <- (p_pos - p_neg) / se_pool

cat("## Hypotheses\n\n")
cat("$$H_0:\\ p_{\\text{positive}} - p_{\\text{negative}} = 0 \\qquad H_a:\\ p_{\\text{positive}} - p_{\\text{negative}} \\neq 0$$\n\n")

cat("## Point estimate\n\n")
cat(sprintf("$\\hat{p}_{\\text{positive}} - \\hat{p}_{\\text{negative}} = %.3f - %.3f = %.3f$\n\n",
            p_pos, p_neg, p_pos - p_neg))

cat("## Test results\n\n")
res3 <- data.frame(
  Quantity = c("Test statistic (z)", "p-value",
               "95% CI lower", "95% CI upper"),
  Value = c(round(z_stat, 3),
            fmt_p(test3$p.value),
            round(test3$conf.int[1], 3),
            round(test3$conf.int[2], 3))
)
print(kable(res3))

cat("\n\n## Interpretation\n\n")
cat(sprintf("- The test statistic %.3f tells us how many standard errors the observed difference in sample proportions lies from zero.\n", z_stat))
cat(sprintf("- The p-value of %s is the probability of seeing a difference at least this extreme if framing had no real effect on recommendations.\n", fmt_p(test3$p.value)))
ci_includes_zero3 <- test3$conf.int[1] <= 0 & test3$conf.int[2] >= 0
cat(sprintf("- We are 95%% confident that the true difference $p_{\\text{positive}} - p_{\\text{negative}}$ lies between %.3f and %.3f. ",
            test3$conf.int[1], test3$conf.int[2]))
if (ci_includes_zero3) {
  cat("Because **0 falls inside** this interval, the data are consistent with framing having no effect on recommendations.\n\n")
} else {
  cat("Because **0 does not fall inside** this interval, we have evidence that the two framings produce different recommendation rates.\n\n")
}
```

\newpage

# Discussion Questions for Students

For each experiment, work through these in groups:

1. **Population vs. sample.** What population are we trying to make inferences about? In what ways might our sample fail to represent that population?
2. **Random assignment vs. random sampling.** Which one(s) did this experiment use, and what does that allow us to conclude?
3. **Hypotheses.** State $H_0$ and $H_a$ in plain English, then in symbols.
4. **Point estimate vs. parameter.** What is the parameter? What is the point estimate? Why are they different?
5. **Test statistic.** Explain what the value means in terms of standard errors.
6. **p-value.** Interpret the p-value in context — *not* as "the probability the null is true."
7. **Confidence interval.** Interpret the 95% CI in context. What does it mean that 0 is (or is not) in the interval?
8. **Conclusion.** At the 5% significance level, do you reject $H_0$? What is your conclusion in the context of the experiment?
